### 2025-03-12 paper/quarto_ecml/paper.tex
Review abstract and introduction

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
% \usepackage{amsthm}
% \theoremstyle{plain}
% \newtheorem{proposition}{Proposition}[section]
% \theoremstyle{plain}
% \newtheorem{corollary}{Research Question}[section]
% \theoremstyle{definition}
% \newtheorem{definition}{Definition}[section]
% \theoremstyle{definition}
% \newtheorem{example}{Example}[section]
% \theoremstyle{remark}
% \AtBeginDocument{\renewcommand*{\proofname}{Proof}}
% \newtheorem*{remark}{Remark}
% \newtheorem*{solution}{Solution}
% \newtheorem{refremark}{Remark}[section]
% \newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{abstract}
We propose a novel training regime called Counterfactual Training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for this desired end goal: Counterfactual Training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning
representations directly from data, guided by narrow objectives such as
predictive accuracy \cite{goodfellow2016deep}. Modern advances in
computing have made it possible to provide such models with ever-growing
degrees of freedom to achieve that task, which has often led them to
outperform traditionally more parsimonious models. Unfortunately, in
doing so, models learn increasingly complex and highly sensitive
representations that we can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{szegedy2013intriguing} showed that artificial neural
networks (ANN) are sensitive to adversarial examples: counterfactuals of
model inputs that yield vastly different model predictions despite being
``imperceptible'' in that they are semantically indifferent from their
factual counterparts. Although some partially effective mitigation
strategies have been proposed, for example \textbf{adversarial training}
\cite{goodfellow2014explaining}, truly robust deep learning (DL) remains
unattainable even for models that are considered shallow by today's
standards \cite{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
Similarly, \cite{zhang2021understanding} showed that state-of-the-art
neural networks are so expressive that they can fit randomly labeled
data. Thus, looking at the predictive performance, the solutions may
seem to provide compelling explanations for the data, when in fact they
are based on purely associative, semantically meaningless patterns. This
poses two related challenges. Firstly, there is no dependable way to
verify if such complex representations correspond to meaningful and
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has occurred in
practice and caused real harm to people who were affected adversely and
often unfairly by automated decision-making (ADM) systems involving
opaque models \cite{oneil2016weapons,mcgregor2021preventing}. Effective
XAI tools can aid us in monitoring models and providing recourse to
individuals to turn adverse outcomes (e.g., ``loan application
rejected'') into positive ones (e.g., ``application accepted'').
\cite{wachter2017counterfactual} propose \textbf{counterfactual
explanations} (CE) as an effective approach to achieve this goal: CEs
explain how factual inputs need to change in order for some fitted model
to produce some desired output, typically involving minimal
perturbations.

To our surprise, the second challenge has not yet attracted any major
consolidated research effort. Specifically, there has been no concerted
effort towards improving improving models' explanatory capacity, which
we will henceforth simply call ``explainability'', defined as the degree
to which learned representations correspond to explanations that are
interpretable and deemed \textbf{plausible} by humans (see
Definition~\ref{def-explainability}). Instead, the choice has typically
been to improve the ability of XAI tools to identify the subset
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with implausible explanations \cite{altmeyer2024faithful}. Fortunately,
recent findings indicate that explainability can arise as byproduct of
regularization techniques aimed at other objectives such as robustness,
generalization, and generative capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.

Building on these findings, we introduce \textbf{counterfactual
training}: a novel training regime explicitly geared towards aligning
model representations with plausible explanations. Our contributions are
as follows:

\begin{itemize}
\tightlist
\item
  We discuss existing related work on improving models and consolidate
  it through the lens of counterfactual explanations
  (Section~\ref{sec-lit}).
\item
  We present our proposed methodological framework that leverages
  faithful counterfactual explanations during the training phase of
  models to achieve the explainability objective
  (Section~\ref{sec-method}).
\item
  Through extensive experiments we demonstrate the counterfactual
  training improve model explainability while maintaining high
  predictive performance. We run ablation studies and grid searches to
  understand how the underlying model components and hyperparameters
  affect outcomes. (Section~\ref{sec-experiments}).
\end{itemize}

Despite some limitations of our approach discussed in
Section~\ref{sec-discussion}, we conclude in
Section~\ref{sec-conclusion} that counterfactual training provides a
practical framework for researchers and practitioners interested in
making opaque models more trustworthy. We also believe that this work
serves as an opportunity for XAI researchers to re-evaluate the trend of
improving XAI tools without improving the underlying models.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search as
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see supplementary appendix.}

@@

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs\footnote{See supplementary appendix for computational details.}.
However, there are three factors that attenuate the impact of this
limitation. First, CT provides counterfactual explanations for the
training samples essentially for free, which may be beneficial in many
ADM systems. Second, we find that CT can retain its value when used as a
``fine-tuning'' training regime for conventionally-trained models.
Third, in principle, CT yields itself to parallel execution, which we
have leveraged for our own experiments.

@@

\textbf{\emph{Interventions on features may have downstream impacts on
fairness.}}\\
- \href{https://arxiv.org/abs/2401.16088}{Equality of opportunity}?\\
- \href{https://proceedings.mlr.press/v202/gao23d/gao23d.pdf}{Social
segregation}?\\
- Supporting AR within one context/system may still unfairly target
recourse-affected individuals within other contexts/systems? In Example
3.1 younger individuals gain access to a loan but their age could still
lead to, e.g., enforcing stronger supervisory mechanisms?

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{abstract}
We propose a novel training regime termed Counterfactual Training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: Counterfactual Training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by \textbf{representation learning}: instead of relying on
features and rules that are carefully hand-crafted by humans, modern
machine learning (ML) models are tasked with learning representations
directly from data, guided by narrow objectives such as predictive
accuracy \cite{goodfellow2016deep}. Modern advances in computing have
made it possible to provide such models with ever-growing degrees of
freedom to achieve that task, which frequently allows them to outperform
traditionally more parsimonious models. Unfortunately, in doing so,
models learn increasingly complex and highly sensitive representations
that humans can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{szegedy2013intriguing} showed that artificial neural
networks (ANN) are sensitive to adversarial examples: perturbed versions
of data instances that yield vastly different model predictions despite
being ``imperceptible'' in that they are semantically indifferent from
their factual counterparts. Even though some partially effective
mitigation strategies have been proposed---most notably
\textbf{adversarial training} \cite{goodfellow2014explaining}---truly
robust deep learning (DL) remains unattainable even for models that are
considered shallow by today's standards \cite{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
Similarly, \cite{zhang2021understanding} showed that state-of-the-art
neural networks are so expressive that they can fit randomly labeled
data. Thus, looking at the predictive performance alone, the solutions
may seem to provide compelling explanations for the data, when in fact
they are based on purely associative, semantically meaningless patterns.
This poses two related challenges. Firstly, there is no dependable way
to verify if such complex representations correspond to meaningful and
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models
\cite{oneil2016weapons,mcgregor2021preventing}. Effective XAI tools can
aid us in monitoring models and providing recourse to individuals to
turn negative outcomes (e.g., ``loan application rejected'') into
positive ones (e.g., ``application accepted''). Our work builds upon
\textbf{counterfactual explanations} (CE) proposed by
\cite{wachter2017counterfactual} as an effective approach to achieve
this goal. CEs prescribe minimal changes for factual inputs that, if
implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major
research interest. Specifically, there has been no concerted effort
towards improving the ``explanatory capacity'' of models, i.e., the
degree to which learned representations correspond to explanations that
are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
Definition~\ref{def-explainability}). Instead, the choice has generally
been to improve the ability of XAI tools to identify the subset of
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with plausible explanations \cite{altmeyer2024faithful}. Fortunately,
recent findings indicate that improved explanatory capacity can arise as
a consequence of regularization techniques aimed at other training
objectives such as robustness, generalization, and generative capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
As further discussed in Section~\ref{sec-lit}, our work consolidates
these findings within a single objective.

\textbf{Specifically, we introduce counterfactual training}: a novel
training regime explicitly meant to align learned representations with
plausible explanations that comply with user requirements. Our
contributions are as follows:

\begin{itemize}
\tightlist
\item
  We present a novel methodological framework that leverages adversarial
  examples and faithful counterfactual explanations during the training
  phase to improve the explanatory capacity and robustness of machine
  learning models (Section~\ref{sec-method}).\\
\item
  We propose a method to enforce global actionability constraints by
  preventing models from assigning importance to immutable features,
  i.e., ones over which decision subjects have no control
  (Section~\ref{sec-method}).\\
\item
  Through extensive experiments we demonstrate that counterfactual
  training promotes explainability while preserving high predictive
  performance. We run ablation studies and grid searches to understand
  how the underlying model components and hyperparameters affect
  outcomes. (Section~\ref{sec-experiments}).
\end{itemize}

Despite some limitations discussed in Section~\ref{sec-discussion}, we
conclude in Section~\ref{sec-conclusion} that counterfactual training
provides a useful framework for researchers and practitioners interested
in making opaque models more trustworthy. We also believe that this work
serves as an opportunity for XAI researchers to re-evaluate the trend of
improving XAI tools without improving the underlying models.

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search as
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see \textbf{?@sec-app-ce}.}

@@

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs (see details in \textbf{?@sec-hardware}). However, there are three
factors that attenuate the impact of this limitation. First, CT provides
counterfactual explanations for the training samples essentially for
free, which may be beneficial in many ADM systems. Second, we find that
CT can retain its value when used as a ``fine-tuning'' training regime
for conventionally-trained models. Third, in principle, CT yields itself
to parallel execution, which we have leveraged for our own experiments.

@@

\textbf{\emph{Interventions on features may have downstream impacts on
fairness.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, counterfactual training could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness} and
potentially reinforcing social segregation \cite{gao2023impact}. Still,
as the referenced publications indicate, such phenomena are not specific
to CT; regrettably, all types of ADM solutions without strong external
protections have been recognized to promote harmful power dynamics
\cite{maas2023machine}.

@@

\newpage{}

### 2025-03-12 paper/sections/intro.qmd
Review abstract and introduction

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such models with ever-growing degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so, models learn increasingly complex and highly sensitive representations that we can no longer easily interpret.

The trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: counterfactuals of model inputs that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts. Although some partially effective mitigation strategies have been proposed, for example **adversarial training** [@goodfellow2014explaining], truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data. Thus, looking at the predictive performance, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations. This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions. On countless occasions, this scenario has occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing]. Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn adverse outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted"). @wachter2017counterfactual propose **counterfactual explanations** (CE) as an effective approach to achieve this goal: CEs explain how factual inputs need to change in order for some fitted model to produce some desired output, typically involving minimal perturbations.

To our surprise, the second challenge has not yet attracted any major consolidated research effort. Specifically, there has been no concerted effort towards improving improving models' explanatory capacity, which we will henceforth simply call "explainability", defined as the degree to which learned representations correspond to explanations that are interpretable and deemed **plausible** by humans (see @def-explainability). Instead, the choice has typically been to improve the ability of XAI tools to identify the subset explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with implausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful].

Building on these findings, we introduce **counterfactual training**: a novel training regime explicitly geared towards aligning model representations with plausible explanations. Our contributions are as follows:

- We discuss existing related work on improving models and consolidate it through the lens of counterfactual explanations (@sec-lit).
- We present our proposed methodological framework that leverages faithful counterfactual explanations during the training phase of models to achieve the explainability objective (@sec-method).
- Through extensive experiments we demonstrate the counterfactual training improve model explainability while maintaining high predictive performance. We run ablation studies and grid searches to understand how the underlying model components and hyperparameters affect outcomes. (@sec-experiments).

Despite some limitations of our approach discussed in @sec-discussion, we conclude in @sec-conclusion that counterfactual training provides a practical framework for researchers and practitioners interested in making opaque models more trustworthy. We also believe that this work serves as an opportunity for XAI researchers to re-evaluate the trend of improving XAI tools without improving the underlying models.

>>>>

Today's prominence of artificial intelligence (AI) has largely been driven by **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such models with ever-growing degrees of freedom to achieve that task, which frequently allows them to outperform traditionally more parsimonious models. Unfortunately, in doing so, models learn increasingly complex and highly sensitive representations that humans can no longer easily interpret.

The trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: perturbed versions of data instances that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts. Even though some partially effective mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data. Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

The first challenge has attracted an abundance of research on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations. Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions. On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing]. Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted"). Our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal. CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.

To our surprise, the second challenge has not yet attracted major research interest. Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see @def-explainability). Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with plausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful]. As further discussed in @sec-lit, our work consolidates these findings within a single objective.

**Specifically, we introduce counterfactual training**: a novel training regime explicitly meant to align learned representations with plausible explanations that comply with user requirements. Our contributions are as follows:

- We present a novel methodological framework that leverages adversarial examples and faithful counterfactual explanations during the training phase to improve the explanatory capacity and robustness of machine learning models (@sec-method).  
- We propose a method to enforce global actionability constraints by preventing models from assigning importance to immutable features, i.e., ones over which decision subjects have no control (@sec-method).  
- Through extensive experiments we demonstrate that counterfactual training promotes explainability while preserving high predictive performance. We run ablation studies and grid searches to understand how the underlying model components and hyperparameters affect outcomes. (@sec-experiments).

Despite some limitations discussed in @sec-discussion, we conclude in @sec-conclusion that counterfactual training provides a useful framework for researchers and practitioners interested in making opaque models more trustworthy. We also believe that this work serves as an opportunity for XAI researchers to re-evaluate the trend of improving XAI tools without improving the underlying models.
